Notes on Perturbational Quantities

### 2009-02-26 docs/pert_notes/pert.tex
updating in progress

%-----------------------------------------------------------------------------
% density
%-----------------------------------------------------------------------------
\section{Density Evolution}
The full density evolution equation is
\begin{equation}
\frac{\partial\rho}{\partial t} = -\nabla\cdot(\rho\Ub) = -\Ub\cdot\nabla\rho 
- \rho\nabla\cdot\Ub. \label{rho equation}
\end{equation}
We predict both $\rho'$ and $\rho_0$ to edges and later reconstruct $\rho$.
The base state density evolution equation is
\begin{equation}
\frac{\partial\rho_0}{\partial t} = -\nabla\cdot(\rho_0 w_0 \eb_r) = 
-w_0\frac{\partial\rho_0}{\partial r} 
\underbrace{-\rho_0\frac{\partial w_0}{\partial r}}_{"\rho_0 ~ \text{force}"}.
\label{rho0 equation}
\end{equation}
Subtract (\ref{rho0 equation}) from (\ref{rho equation}) and rearrange
terms, noting that $\Ub = \Ubt + w_o\eb_r$, to obtain the
perturbational density equation,
\begin{equation}
\frac{\partial\rho'}{\partial t} = -\Ub\cdot\nabla\rho' \underbrace{- \rho'\nabla\cdot\Ub 
- \nabla\cdot(\rho_0\Ubt)}_{"\rho' ~ \text{force}"}. 
\label{rhoprime equation}
\end{equation}

\subsection{Predicting $\rho'$ at edges}
We define $\rho' = \rho^n - \rho_0^n$.  Then we predict $\rho'$ to
edges using {\tt make\_edge\_scal} in {\tt scalar\_advance} and the
underbraced term in (\ref{rhoprime equation}) as the forcing.

\subsection{Predicting $\rho_0$ at edges}\label{Predicting rho0 at edges}
There are two ways to predict $\rho_0$ at edges.
\begin{enumerate}
\item We call {\tt make\_edge\_state\_1d} using the underbraced term
in (\ref{rho0 equation}) as the forcing.  This gives us
$\rho_0^{n+\half,{\rm pred}}$.  This term is used to advect $\rho_0$
in {\tt advect\_base} and therefore must also be used to compute
$\eta_\rho$, which will be used in {\tt correct\_base}.
\item We define $\rho_0^{n+\half,{\rm avg}} = (\rho_0^n +
\rho_0^{(2)})/2$.  We compute $\rho_0^{(2)}$ from $\rho_0^n$ using
{\tt advect\_base}, which advances equation (\ref{rho0 equation})
through $\Delta t$ in time.  The $(2)$ in the superscript indicates
that we have not called {\tt correct\_base} yet, which computes
$\rho_0^{n+1}$ from $\rho_0^{(2)}$.  We use $\rho_0^{(2)}$ rather than
$\rho_0^{n+1}$ to construct $\rho_0^{n+\half,{\rm avg}}$ to be
consistent with the fact that we do not include an $\eta_\rho$
contribution to compute $\rho'$ at edges.  $\rho_0^{n+\half,{\rm
avg}}$ is used to construct $\rho$ at edges from $\rho'$ at edges, and
this $\rho$ at edges is used to compute fluxes for $\rho X_k$.
\end{enumerate}

@@

\subsection{Predicting $(\rho h)'$ ({\tt enthalpy\_pred\_type = 1} - default)}
Due to operator splitting, the call to {\tt react\_state} has already
been made.  Hence, the goal is to compute an edge state enthalpy
starting from $(\rho h)^{(1)}$ using an enthalpy equation that does
not include the $\rho H_{\rm nuc}$ and $\rho H_{\rm ext}$ terms, where
were already accounted for in {\tt react\_state}
\begin{equation}
\frac{\partial(\rho h)}{\partial t} = -\Ub\cdot\nabla(\rho h) - \rho h\nabla\cdot\Ub 
+ \psi + (\Ubt \cdot \er) \frac{\partial p_0}{\partial r} + \nabla\cdot\kth\nabla T. \label{rhoh equation}
\end{equation}
We define the base state enthalpy evolution equation as
\begin{eqnarray}
\frac{\partial(\rho h)_0}{\partial t} &=& -\nabla\cdot[(\rho h)_0 w_0\eb_r] 
+ \frac{D_0p_0}{Dt} \nonumber \\
&=& -w_0\frac{\partial(\rho h)_0}{\partial r} 
- \underbrace{(\rho h)_0\frac{\partial w_0}{\partial r}+ \psi}_{"(\rho h)_0 ~ \text{force}"}
\enskip .\label{rhoh0 equation}
\end{eqnarray}
Subtracting (\ref{rhoh0 equation}) from (\ref{rhoh equation}) and rearranging terms gives 
the perturbational enthalpy equation
\begin{eqnarray}
\frac{\partial(\rho h)'}{\partial t} &=& -\nabla\cdot[(\rho h)'\Ub] 
- \nabla\cdot[(\rho h)_0\Ubt] + (\Ubt \cdot \er)\frac{\partial p_0}{\partial r} 
+ \nabla\cdot\kth\nabla T\nonumber \\
&=& -\Ub\cdot\nabla(\rho h)' \underbrace{- (\rho h)'\nabla\cdot\Ub 
- \nabla\cdot[(\rho h)_0\Ubt] + (\Ubt \cdot \er)\frac{\partial p_0}{\partial r}
+ \nabla\cdot\kth\nabla T}_{"(\rho h)' ~ \text{force}"}. \label{rhohprime equation}
\end{eqnarray}

\subsubsection{Advancing $(\rho h)_0$}
\begin{itemize}
\item {\bf Using full\_rhoh0\_evolution = F:}\\
 Define $(\rho h)_0^{(1)} = \overline{(\rho h)^{(1)}}$  
and $(\rho h)' = (\rho h)^{(1)}-(\rho h)_0^{(1)}$.  Then, $(\rho h)_0^{(1)} \rightarrow$ 
{\tt advect\_base} $\rightarrow (\rho h)_0^{(2)}$.
\item {\bf Using full\_rhoh0\_evolution = T:}\\
 Instead of defining $(\rho h)_0^{(1)} = \overline{(\rho h)^{(1)}}$, the code carries 
through the complete evolution of $(\rho h)_0$ using $(\rho h)_0^n \rightarrow$ 
{\tt react\_base} $\rightarrow (\rho h)_0^{(1)} \rightarrow$ {\tt advect\_base}
$\rightarrow (\rho h)_0^{(2)} \rightarrow$ {\tt react\_base} $\rightarrow (\rho h)_0^{n+1}$.  
There is no {\tt correct\_base} function for $(\rho h)_0$ written yet, nor is any 
adjustment made to sync up $(\rho h)_0$ after the thermal diffusion step.
\end{itemize}

@@

\subsection{Predicting $h$ ({\tt enthalpy\_pred\_type = 2})}

\subsection{Predicting $T$ then $(\rho h)'$ ({\tt enthalpy\_pred\_type = 3})}

\subsection{Predicting $T$ then $h$ ({\tt enthalpy\_pred\_type = 4})}

>>>>

%-----------------------------------------------------------------------------
% density
%-----------------------------------------------------------------------------
\section{Density Evolution}
The full density evolution equation is
\begin{eqnarray}
\frac{\partial\rho}{\partial t} &=& -\nabla\cdot(\rho\Ub) \nonumber \\
&=& -\Ub\cdot\nabla\rho - \rho\nabla\cdot\Ub. \label{rho equation}
\end{eqnarray}
We predict both $\rho'$ and $\rho_0$ to edges separately and later use them to 
reconstruct $\rho$ at edges.  The base state density evolution equation is
\begin{equation}
\frac{\partial\rho_0}{\partial t} = -\nabla\cdot(\rho_0 w_0 \eb_r) = 
-w_0\frac{\partial\rho_0}{\partial r} 
\underbrace{-\rho_0\frac{\partial w_0}{\partial r}}_{"\rho_0 ~ \text{force}"}.
\label{rho0 equation}
\end{equation}
Subtract (\ref{rho0 equation}) from (\ref{rho equation}) and rearrange
terms, noting that $\Ub = \Ubt + w_o\eb_r$, to obtain the
perturbational density equation,
\begin{equation}
\frac{\partial\rho'}{\partial t} = -\Ub\cdot\nabla\rho' \underbrace{- \rho'\nabla\cdot\Ub 
- \nabla\cdot(\rho_0\Ubt)}_{f_{\rho'}},
\label{rhoprime equation}
\end{equation}
where the underbraced term is the "$\rho'$ force'', which is computed 
in {\tt modify\_scal\_force}.

\subsection{Predicting $\rho'$ at edges}
We define $\rho' = \rho^{(1)} - \rho_0^n$.  Then we predict $\rho'$ to
edges using {\tt make\_edge\_scal} in {\tt scalar\_advance} and the
underbraced term in (\ref{rhoprime equation}) as the forcing.

\subsection{Predicting $\rho_0$ at edges}\label{Predicting rho0 at edges}
There are two ways to predict $\rho_0$ at edges.
\begin{enumerate}

\item We call {\tt make\_edge\_state\_1d} using the underbraced term
in (\ref{rho0 equation}) as the forcing.  This gives us
$\rho_0^{n+\half,{\rm pred}}$.  This term is used to advect $\rho_0$
in {\bf Advect Base Density}.  In plane-parallel geometries, we also use
$\rho_0^{n+\half,{\rm pred}}$ to compute $\etarho$, which will be used 
to compute $\psi$.

\item We define $\rho_0^{n+\half,{\rm avg}} = (\rho_0^n +
\rho_0^{(2)})/2$.  We compute $\rho_0^{(2)}$ from $\rho_0^n$ using
{\bf Advect Base Density}, which advances equation (\ref{rho0 equation})
through $\Delta t$ in time.  The $(2)$ in the superscript indicates
that we have not called {\bf Correct Base} yet, which computes
$\rho_0^{n+1}$ from $\rho_0^{(2)}$.  We use $\rho_0^{(2)}$ rather than
$\rho_0^{n+1}$ to construct $\rho_0^{n+\half,{\rm avg}}$ to be
consistent with the fact that we do not include an $\eta_\rho$
contribution to compute $\rho'$ at edges.  $\rho_0^{n+\half,{\rm
avg}}$ is used to construct $\rho$ at edges from $\rho'$ at edges, and
this $\rho$ at edges is used to compute fluxes for $\rho X_k$.
\end{enumerate}

@@

\subsection{Predicting $(\rho h)'$ ({\tt enthalpy\_pred\_type = 1} - default)}
Due to operator splitting, the call to {\tt react\_state} has already
been made.  Hence, the goal is to compute an edge state enthalpy
starting from $(\rho h)^{(1)}$ using an enthalpy equation that does
not include the $\rho H_{\rm nuc}$ and $\rho H_{\rm ext}$ terms, where
were already accounted for in {\tt react\_state}
\begin{equation}
\frac{\partial(\rho h)}{\partial t} = -\Ub\cdot\nabla(\rho h) - \rho h\nabla\cdot\Ub 
+ \psi + (\Ubt \cdot \er) \frac{\partial p_0}{\partial r} + \nabla\cdot\kth\nabla T. \label{rhoh equation}
\end{equation}
We define the base state enthalpy evolution equation as
\begin{eqnarray}
\frac{\partial(\rho h)_0}{\partial t} &=& -\nabla\cdot[(\rho h)_0 w_0\eb_r] 
+ \frac{D_0p_0}{Dt} \nonumber \\
&=& -w_0\frac{\partial(\rho h)_0}{\partial r} 
- \underbrace{(\rho h)_0\frac{\partial w_0}{\partial r}+ \psi}_{"(\rho h)_0 ~ \text{force}"}
\enskip .\label{rhoh0 equation}
\end{eqnarray}
Subtracting (\ref{rhoh0 equation}) from (\ref{rhoh equation}) and rearranging terms gives 
the perturbational enthalpy equation
\begin{eqnarray}
\frac{\partial(\rho h)'}{\partial t} &=& -\nabla\cdot[(\rho h)'\Ub] 
- \nabla\cdot[(\rho h)_0\Ubt] + (\Ubt \cdot \er)\frac{\partial p_0}{\partial r} 
+ \nabla\cdot\kth\nabla T\nonumber \\
&=& -\Ub\cdot\nabla(\rho h)' \underbrace{- (\rho h)'\nabla\cdot\Ub 
- \nabla\cdot[(\rho h)_0\Ubt] + (\Ubt \cdot \er)\frac{\partial p_0}{\partial r}
+ \nabla\cdot\kth\nabla T}_{"(\rho h)' ~ \text{force}"}. \label{rhohprime equation}
\end{eqnarray}
Note that the pressure term in the $(\rho h)'$ force can be represented as:
\begin{equation}
(\Ubt \cdot \er)\frac{\partial p_0}{\partial r} \equiv \Ubt\cdot\nabla p_0 = 
\nabla\cdot(\Ubt p_0) - p_0\nabla\cdot\Ubt.
\end{equation}
For spherical problems, we use the RHS representation of this term, as we have
seen it gives better results.

\subsubsection{Advancing $(\rho h)_0$}
Define $(\rho h)_0^{n} = \overline{(\rho h)^{(1)}}$  
and $(\rho h)' = (\rho h)^{(1)}-(\rho h)_0^{n}$.  Then, $(\rho h)_0^{n} \rightarrow$ 
{\bf Advect Base Enthalpy} $\rightarrow (\rho h)_0^{n+1}$.

@@
